---
title: "Good-238 tuned amphora tree — 76 PC"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 9
    fig_height: 6
params:
  # Good-238 post-QC types, tuned FBD, 76 SRVF PCs. Sibling of
  # three_clade_plots_good238_38_tuned.Rmd (identical but for pc_tag) and of
  # three_clade_plots_all297.Rmd. pc_tag drives every derived path below; leave
  # it at "76" in this file. See section 0 -- the 38 and 76 trees are NOT equally
  # trustworthy, and the difference is not visible in the ESS table.
  pc_tag: "76"
  repo_dir: "/Users/armandleroi/Documents/Amphora_phylogeny/phylogenies/2026_07_three_clade_tree"
  data_dir: "/Users/armandleroi/Documents/Amphora_phylogeny/Making amphora types/05_2026_amphora_297_types"
  profile_file: "2026_amphora_297_types_profiles_with_metadata.csv"
  selected_file: "selected_taxa_all297.csv"   # superset; filtered to the 238 tips below
  taxa_file: "data/mini_taxa_good238.tsv"
  bp_zero: 1950
  pot_offset: 0.03
  pot_height: 0.85
  pot_stretch: 1.0
  fig_w: 13
  fig_h: 52
  burnin: 0.10
  support_cutoff: 0.50
  sigma2_floor: 0.001      # lower bound of dnLoguniform(1e-3, 1) in scripts/strict_clock.Rev
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ape); library(treeio); library(ggtree); library(ggplot2)
library(dplyr); library(tidyr); library(data.table)
options(ignore.negative.edge = TRUE)
tag      <- params$pc_tag
run_dir  <- file.path(params$repo_dir, sprintf("runs/good238_%s_tuned/output", tag))
runs_dir <- file.path(params$repo_dir, "runs")
# NB: the salvage MCC (built from the three _run_N.trees after the walltime kill),
# not the clean-finish *_mcc.tree, which mcmcmc never got to write.
tree_file  <- sprintf("simple_run_good238_%s_tuned_mcc_perrun.tree", tag)
pc_file    <- sprintf("data/mini_character_good238_srvf_%s.nex", tag)
log_prefix <- sprintf("simple_run_good238_%s_tuned", tag)
```

## 0. Run status — read this before interpreting anything

```{r status}
cat(sprintf("Tree: good238 SRVF-%s PC, tuned FBD (FBD_tuned.Rev), 238 post-QC types.\n\n", tag))
cat("Both runs were killed by the 72h walltime at 93-96%% of their 2,000,000\n")
cat("generations, after a 50,000-generation burnin. That truncation is immaterial.\n\n")
cat("The FBD machinery converged well in BOTH runs:\n")
cat("  psi off zero (~0.006, matching the value the data imply) -- the corrected\n")
cat("  dnExponential(100) prior worked; diversification/turnover ESS ~10-20k;\n")
cat("  origin_time clear of its 4000 BP ceiling.\n")
cat("  BUT cor(diversification, turnover) ~ -0.93 to -0.95: the lambda=mu ridge was\n")
cat("  rotated into well-sampling coordinates, NOT removed. Quote net diversification\n")
cat("  and turnover; never lambda and mu separately (they are ':=' deterministic).\n\n")
if (tag == "76") {
  cat("*** THIS (76 PC) TREE IS COMPROMISED ***\n")
  cat("sigma2 is pinned against the 1e-3 floor of dnLoguniform(1e-3, 1): ~90%% of the\n")
  cat("posterior lies within 1%% of the bound, across only ~312 distinct values. The BM\n")
  cat("rate is set by the PRIOR, not estimated from the data -- PCs 39-76 carry little\n")
  cat("variance and drag the single shared rate under the floor. Section 10 quantifies\n")
  cat("this. Treat branch-length/rate statements here as unsupported.\n\n")
  cat("NOTE the trap: 76 PC sigma2 shows ESS ~10,700 and PSRF ~1.01, which LOOKS like\n")
  cat("excellent mixing. A parameter welded to a boundary is near-constant, and\n")
  cat("near-constant parameters yield flattering ESS/PSRF. Do not read that as health.\n")
} else {
  cat("This (38 PC) tree is the one to use. sigma2 is low but genuinely estimated and\n")
  cat("clear of the floor (min 0.00112 vs floor 0.001, 4,791 distinct values). Its\n")
  cat("ESS ~70 / PSRF ~1.08 is a real caveat -- slow mixing in that one parameter --\n")
  cat("but it is a lesser problem than the 76 PC truncation, and it does not undermine\n")
  cat("topology or chronology, which rest on parameters that sampled well.\n")
}
```

## 1. Read the tree

```{r read}
tr  <- read.beast(file.path(run_dir, tree_file))
phy <- as.phylo(tr)
tr
```

## 2. Tip metadata (clade = shape class, origin = Greek/Italiote)

```{r meta}
# good238 is the post-QC subset of the 297, so the 297 selection table is the
# metadata source; the join is filtered to whatever tips this tree actually has.
sel <- read.csv(file.path(params$repo_dir, params$selected_file), stringsAsFactors = FALSE)
stopifnot("tips missing from selected_taxa" = all(phy$tip.label %in% sel$type))
meta <- tibble::tibble(label = phy$tip.label) |>
  left_join(sel |> transmute(label = type, shape_general, shape, place,
              year = round((date_earliest + date_latest) / 2)), by = "label") |>
  mutate(clade = case_when(
           shape_general == "pelike"        ~ "pelike",
           shape_general == "loutrophoros"  ~ "loutrophoros",
           grepl("panathenaic", shape)      ~ "panathenaic",
           grepl("neck", shape)             ~ "neck",
           grepl("belly", shape)            ~ "belly",
           TRUE                             ~ "amphora (other)"),
         origin = ifelse(place == "greek_italiote", "Italiote", "Greek"),
         id = sub(".*_([0-9]+)$", "\\1", label),
         tip_label = sprintf("%s %s %d BCE (%s)", origin, clade, -year, id))
stopifnot(!any(is.na(meta$clade)))
pal_clade <- c(neck = "grey45", belly = "#B2182B", `amphora (other)` = "grey70",
               pelike = "#E69F00", panathenaic = "#56B4E9", loutrophoros = "#009E73")
knitr::kable(meta |> count(clade, origin) |> pivot_wider(names_from = origin, values_from = n, values_fill = 0))
```

## 3. Time axis (absolute chronology)

MCC branch lengths are relative (youngest tip at 0); we anchor on the youngest
tip's age from `mini_taxa_good238.tsv` and read the axis in calendar years.
(Anchor uncertainty is not propagated here — it narrows intervals slightly.)

```{r axis}
tax <- read.delim(file.path(params$repo_dir, params$taxa_file))
depth       <- ape::node.depth.edgelength(phy)
root_offset <- max(depth)
rel_age     <- root_offset - depth
anchor_bp   <- min((tax$min_age + tax$max_age) / 2)     # youngest tip (BP)
node_year   <- (params$bp_zero - anchor_bp) - rel_age
root_year   <- node_year[ape::Ntip(phy) + 1]
cat(sprintf("Youngest tip ~ %.0f BCE ; Root ~ %.0f BCE (tree height %.0f yr)\n",
            params$bp_zero - anchor_bp, -root_year, root_offset))
p <- ggtree(tr, size = 0.3) %<+% meta
time_axis <- function(headroom = 0.55, left = 0) {
  scale_x_continuous(
    limits = c(min(0, left), root_offset * (1 + headroom)),
    breaks = scales::breaks_pretty(6)(c(min(0, left), root_offset)),
    labels = function(x) { yr <- round(x + root_year)
      ifelse(yr < 0, paste0(abs(yr), " BCE"), ifelse(yr == 0, "0", paste0(yr, " CE"))) })
}
```

## 4. Plot 1 — MCC tree, node labels = posterior support

```{r plot1, fig.width=params$fig_w, fig.height=params$fig_h}
p +
  geom_tiplab(aes(label = tip_label, colour = clade), size = 1.1, offset = 6, show.legend = FALSE) +
  geom_tippoint(aes(colour = clade, shape = origin), size = 1.4) +
  geom_text2(aes(subset = !isTip & !is.na(posterior) & posterior >= params$support_cutoff,
                 label = sprintf("%.2f", posterior)), hjust = 1.2, vjust = -0.5,
             size = 1.3, colour = "grey30") +
  scale_colour_manual(values = pal_clade, name = "Clade") +
  scale_shape_manual(values = c(Greek = 16, Italiote = 17), name = "Origin") +
  theme_tree2() + coord_cartesian(clip = "off") +
  labs(x = NULL, title = sprintf("good238 SRVF-%s tuned: MCC, node labels = posterior (>= %.2f)", tag, params$support_cutoff)) +
  time_axis()
```

## 5. Plot 2 — 95% HPD on node ages (needs the thinned posterior trees)

The six `*_run_N.trees` posteriors (~175 MB each) were deliberately left on the
HPC by `pull_good238_tuned.sh`. This section runs only if thinned copies are
present locally; otherwise it skips and every later section still works.

To enable it: `sh pull_good238_tuned.sh --trees`, then thin to
`runs/thin_good238_<tag>_tuned_run_<r>.trees`.

```{r abs-hpd}
thin_files <- file.path(runs_dir, sprintf("thin_good238_%s_tuned_run_%d.trees", tag, 1:3))
have_thin  <- all(file.exists(thin_files))
if (!have_thin) {
  cat("Thinned posterior trees not found locally — skipping the HPD section.\n")
  cat("Missing:\n"); cat(paste(" ", basename(thin_files), collapse = "\n"), "\n")
} else {
  read_thin <- function(f) {
    L <- readLines(f)[-1]
    read.tree(text = gsub("\\[[^]]*\\]", "", sub("^([^\t]*\t){4}", "", L)))
  }
  trees <- do.call(c, lapply(thin_files, read_thin))
  b <- floor(length(trees) * params$burnin); trees <- trees[(b + 1):length(trees)]
  canon <- sort(phy$tip.label)
  ckey  <- function(t, ix) paste(sort(match(t$tip.label[ix], canon)), collapse = ",")
  acc <- new.env(hash = TRUE, parent = emptyenv())
  for (ts in trees) {
    dd <- node.depth.edgelength(ts); pp <- prop.part(ts)
    ages <- (max(dd) - dd)[ape::Ntip(ts) + seq_along(pp)] + anchor_bp   # fixed anchor
    for (i in seq_along(pp)) { k <- ckey(ts, pp[[i]]); acc[[k]] <- c(acc[[k]], ages[i]) }
  }
  hdi <- function(x, prob = 0.95) { x <- sort(x); n <- length(x); m <- max(1, floor(n * prob))
    i <- which.min(x[m:n] - x[1:(n - m + 1)]); c(x[i], x[i + m - 1]) }
  ntip <- ape::Ntip(phy); mcc_pp <- prop.part(phy)
  node_hpd <- lapply(seq_along(mcc_pp), function(i) {
    a <- acc[[ckey(phy, mcc_pp[[i]])]]; if (is.null(a) || length(a) < 2) return(NULL)
    tibble::tibble(node = ntip + i, freq = length(a) / length(trees),
                   med = median(a), lo = hdi(a)[1], hi = hdi(a)[2]) }) |> bind_rows()
}
```

```{r plot2, fig.width=params$fig_w, fig.height=params$fig_h, eval=have_thin}
bp_to_x <- function(bp) (params$bp_zero - bp) - root_year
bars <- node_hpd |> left_join(p$data |> select(node, y), by = "node") |>
  mutate(x_lo = bp_to_x(hi), x_hi = bp_to_x(lo), x_med = bp_to_x(med))
p +
  geom_segment(data = bars, aes(x = x_lo, xend = x_hi, y = y, yend = y, alpha = freq),
               linewidth = 1.5, colour = "#377EB8", lineend = "round") +
  geom_point(data = bars, aes(x = x_med, y = y), size = 0.5, colour = "grey20") +
  geom_tiplab(aes(label = tip_label, colour = clade), size = 1.0, offset = 6, show.legend = FALSE) +
  scale_colour_manual(values = pal_clade, guide = "none") +
  scale_alpha_continuous(range = c(0.15, 0.9), name = "clade freq") +
  theme_tree2() + coord_cartesian(clip = "off") +
  labs(title = "95% HPD on node ages; opacity = clade frequency in the posterior") +
  time_axis(headroom = 0.75, left = min(bars$x_lo, 0))
```

The MCC tree itself still carries per-node `age_95%_HPD` annotations from
RevBayes, so node-age uncertainty is available even without this section:

```{r mcc-hpd-summary}
# treeio renames the NEXUS "age_95%_HPD" annotation (it appears as age_0.95_HPD),
# so find the column rather than hardcoding a name that varies by version.
d   <- as_tibble(tr)
hpd <- grep("HPD", names(d), value = TRUE)[1]
if (is.na(hpd)) {
  cat("No node-age HPD annotation found on this tree.\n")
} else {
  v <- d[[hpd]]
  ok <- vapply(v, function(z) length(unlist(z)) == 2 && all(is.finite(as.numeric(unlist(z)))), logical(1))
  w  <- vapply(v[ok], function(z) diff(range(as.numeric(unlist(z)))), numeric(1))
  cat(sprintf("column '%s': %d annotated nodes\n", hpd, sum(ok)))
  cat(sprintf("95%% HPD width on node ages: median %.0f yr, IQR %.0f-%.0f yr, max %.0f yr\n",
              median(w), quantile(w, .25), quantile(w, .75), max(w)))
}
```

## 6. Plot 3 — support as node size

```{r plot3, fig.width=params$fig_w, fig.height=params$fig_h}
p +
  geom_nodepoint(aes(size = as.numeric(posterior)), colour = "grey25", alpha = 0.7, na.rm = TRUE) +
  geom_tippoint(aes(colour = clade, shape = origin), size = 1.3) +
  geom_tiplab(aes(label = tip_label, colour = clade), size = 1.0, offset = 6, show.legend = FALSE) +
  scale_size_continuous(range = c(0.2, 3.5), name = "Posterior", limits = c(0, 1)) +
  scale_colour_manual(values = pal_clade, name = "Clade") +
  scale_shape_manual(values = c(Greek = 16, Italiote = 17), name = "Origin") +
  theme_tree2() + coord_cartesian(clip = "off") +
  labs(title = "Clade support (node-point size)") + time_axis()
```

```{r support-dist, fig.width=8, fig.height=3.2}
sup <- as_tibble(tr) |> filter(!is.na(posterior)) |> mutate(posterior = as.numeric(posterior))
cat(sprintf("internal nodes with posterior >= 0.95: %d / %d (%.1f%%);  >= 0.50: %d (%.1f%%)\n",
            sum(sup$posterior >= .95), nrow(sup), 100*mean(sup$posterior >= .95),
            sum(sup$posterior >= .50), 100*mean(sup$posterior >= .50)))
ggplot(sup, aes(posterior)) +
  geom_histogram(bins = 40, fill = "#377EB8", colour = "white", linewidth = .2) +
  geom_vline(xintercept = c(.5, .95), linetype = 2, colour = "grey40") +
  theme_bw() + labs(title = sprintf("good238 SRVF-%s: distribution of node support", tag),
                    x = "posterior probability", y = "nodes")
```

## 7. Plot 4 — pot profiles at the tips, filled by clade

```{r profiles}
prof_all <- read.csv(file.path(params$data_dir, params$profile_file))
stopifnot(all(phy$tip.label %in% prof_all$type))
tipxy <- p$data |> filter(isTip) |> select(label, tx = x, ty = y)
headroom <- 0.75
k  <- (root_offset * (1 + headroom) / ape::Ntip(phy)) * (params$fig_h / params$fig_w)
sy <- params$pot_height / 3; sx <- sy * k * params$pot_stretch
pots <- prof_all |> filter(type %in% phy$tip.label) |>
  select(label = type, point_order, px = x, py = y) |>
  left_join(tipxy, by = "label") |> left_join(select(meta, label, clade), by = "label") |>
  arrange(label, point_order) |>
  mutate(x = tx + params$pot_offset * root_offset + px * sx, y = ty + py * sy)
pot_span <- max(abs(prof_all$x)) * sx
```

```{r plot4, fig.width=params$fig_w, fig.height=params$fig_h}
# darker label palette (the amber pelike #E69F00 is barely legible on white)
pal_label <- pal_clade; pal_label["pelike"] <- "#8a6a00"
tiplab <- distinct(pots, label, tx, ty) |> left_join(meta, by = "label") |>
  mutate(txt = sprintf("%s | %s | %d BCE | #%s", clade, origin, -year, id))
p4 <- p +
  # period-transition guides (relative-x = calendar year - root_year)
  geom_vline(xintercept = c(-480 - root_year, -330 - root_year),
             linetype = "dotted", colour = "grey55", linewidth = 0.4) +
  annotate("text", x = c(-480 - root_year, -330 - root_year), y = ape::Ntip(phy) + 2,
           label = c("480 BCE  Archaic/Classical", "330 BCE  Classical/Hellenistic"),
           size = 2.3, colour = "grey45", hjust = 0, vjust = 0) +
  geom_polygon(data = pots, aes(x = x, y = y, group = label, fill = clade),
               colour = "grey25", linewidth = 0.15) +
  geom_tippoint(size = 0.6, colour = "grey30") +
  # ALL internal-node posterior supports (not just >= cutoff)
  geom_text2(aes(subset = !isTip & !is.na(posterior),
                 label = sprintf("%.2f", as.numeric(posterior))),
             size = 1.3, colour = "grey35", hjust = 1.15, vjust = -0.4) +
  # per-tip label: class | origin | date | #id, coloured by clade
  geom_text(data = tiplab,
            aes(x = tx + params$pot_offset * root_offset + 1.35 * pot_span, y = ty,
                label = txt, colour = clade),
            hjust = 0, size = 1.9, show.legend = FALSE) +
  scale_fill_manual(values = pal_clade, name = "Clade") +
  scale_colour_manual(values = pal_label, guide = "none") +
  theme_tree2() + coord_cartesian(clip = "off") +
  theme(plot.margin = margin(4, 430, 4, 4)) +
  labs(title = sprintf("good238 SRVF-%s tuned: profiles + all node supports + labels (class | origin | date | #id)", tag)) +
  time_axis(headroom = headroom)
# high-resolution vector PDF (zoom as far as needed to read labels)
ggsave(file.path(run_dir, sprintf("tree_profiles_labelled_good238_%s_tuned.pdf", tag)),
       p4, device = cairo_pdf, width = 22, height = params$fig_h, limitsize = FALSE)
p4
```

## 8. Plot 5 — pots filled by measured shape (PC1), outlined by clade

```{r read-pcs}
parse_cont_nex <- function(f) {
  L <- readLines(f); L <- L[grepl("^[[:space:]]+[^[:space:]]+[[:space:]]+-?[0-9]", L)]
  parts <- strsplit(trimws(L), "[[:space:]]+"); tx <- vapply(parts, `[`, "", 1)
  vals <- lapply(parts, function(p) as.numeric(p[-1])); ord <- unique(tx)
  out <- t(vapply(ord, function(t) unlist(vals[tx == t]),
                  numeric(length(unlist(vals[tx == ord[1]])))))
  colnames(out) <- paste0("PC", seq_len(ncol(out))); rownames(out) <- ord; out
}
pcs  <- parse_cont_nex(file.path(params$repo_dir, pc_file))
pcdf <- tibble::as_tibble(pcs[, 1:min(5, ncol(pcs))], rownames = "label")
aspect <- sapply(pcdf$label, function(t) {
  s <- filter(prof_all, type == t); (max(s$y) - min(s$y)) / (max(s$x) - min(s$x)) })
r_pc1 <- cor(pcdf$PC1, aspect)
cat(sprintf("cor(PC1, height/width) = %.2f\n", r_pc1))
if (r_pc1 < 0) { pcs[, "PC1"] <- -pcs[, "PC1"]; pcdf$PC1 <- -pcdf$PC1 }
```

```{r plot5, fig.width=params$fig_w, fig.height=params$fig_h}
pots_pc <- pots |> left_join(pcdf |> select(label, PC1), by = "label")
p +
  geom_polygon(data = pots_pc, aes(x = x, y = y, group = label, fill = PC1, colour = clade),
               linewidth = 0.4) +
  scale_fill_gradient2(low = "#56B4E9", mid = "grey93", high = "#E69F00", midpoint = 0,
                       name = "PC1\n(squat -> slender)") +
  scale_colour_manual(values = pal_clade, name = "Clade") +
  guides(colour = guide_legend(override.aes = list(fill = NA, linewidth = 1.2))) +
  theme_tree2() + coord_cartesian(clip = "off") +
  labs(title = "Fill = measured shape (PC1); outline = clade") +
  time_axis(headroom = headroom)
```

```{r morphospace, fig.width=8, fig.height=7}
msp <- pcdf |> left_join(meta, by = "label")
ggplot(msp, aes(PC1, PC2)) +
  geom_hline(yintercept = 0, colour = "grey85") + geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(aes(fill = clade, shape = origin), size = 2.4, colour = "grey25", stroke = 0.3) +
  scale_fill_manual(values = pal_clade, name = "Clade") +
  scale_shape_manual(values = c(Greek = 21, Italiote = 24), name = "Origin") +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  theme_bw() + labs(title = sprintf("Morphospace of the good-238 types, SRVF-%s (PC1 = squat<->slender)", tag))
```

## 9. Posterior parameters & traces

The tuned model samples `diversification` and `turnover`; `speciation_rate` and
`extinction_rate` are deterministic (`:=`) functions of them and are logged but
must not be interpreted individually.

```{r params-post}
logs <- list.files(run_dir, pattern = sprintf("^%s_run_[0-9]+\\.log$", log_prefix), full.names = TRUE)
stopifnot("no run logs found" = length(logs) > 0)
sampled <- c("diversification", "turnover", "psi", "sigma2", "origin_time")
derived <- c("speciation_rate", "extinction_rate")
post <- rbindlist(lapply(seq_along(logs), function(i) {
  d <- fread(logs[i], select = c("Iteration", sampled, derived))
  d <- d[(floor(.N * params$burnin) + 1):.N]; d$run <- factor(i); d }))
# chains stopped at different generations (walltime kill) -> common window
common <- post[, min(max(Iteration)), by = run][, min(V1)]
post <- post[Iteration <= common]
cat(sprintf("common generation window: 0-%s (%.1f%% of the 2,000,000 target)\n",
            format(common, big.mark = ","), common / 2e6 * 100))
post |> select(all_of(c(sampled, derived))) |> pivot_longer(everything()) |> group_by(name) |>
  summarise(median = median(value), lower = quantile(value, .025),
            upper = quantile(value, .975)) |>
  mutate(interpret = ifelse(name %in% derived, "DETERMINISTIC — do not report", "sampled")) |>
  knitr::kable(digits = 5)
```

```{r trace, fig.height=6, fig.width=10}
post |> select(Iteration, run, all_of(sampled)) |> pivot_longer(all_of(sampled)) |>
  ggplot(aes(Iteration, value, colour = run)) +
  geom_line(alpha = 0.6, linewidth = 0.25) + facet_wrap(~name, scales = "free_y") +
  scale_colour_brewer(palette = "Dark2") + theme_bw() +
  labs(title = sprintf("good238 SRVF-%s tuned: traces of the sampled parameters", tag), y = NULL)
```

## 10. MCMC diagnostics (ESS / PSRF) and the sigma2 boundary check

```{r ess}
library(coda)
allp <- c("origin_time", "sigma2", "diversification", "turnover", "psi")
runs <- split(post, post$run)
nmin <- min(sapply(runs, nrow))
runs <- lapply(runs, function(d) d[seq_len(nmin)])
ess  <- sapply(runs, function(d) sapply(d[, ..allp], \(x) coda::effectiveSize(as.mcmc(x))))
psrf <- coda::gelman.diag(coda::mcmc.list(lapply(runs, \(d) as.mcmc(as.matrix(d[, ..allp])))),
                          multivariate = FALSE, autoburnin = FALSE)$psrf
tibble::tibble(parameter = allp, run1 = round(ess[, 1]), run2 = round(ess[, 2]),
               run3 = round(ess[, 3]), psrf = round(psrf[allp, 1], 3)) |>
  arrange(pmin(run1, run2, run3)) |> knitr::kable()
cat(sprintf("\ncor(diversification, turnover) = %.4f  (ridge rotated, not removed)\n",
            cor(post$diversification, post$turnover)))
cat(sprintf("cor(lambda, mu) = %.4f  (deterministic; quote net-div/turnover)\n",
            cor(post$speciation_rate, post$extinction_rate)))
```

**The boundary check that the ESS table cannot show.** A parameter welded to a
prior bound is near-constant, and near-constant parameters return excellent ESS
and PSRF. So test the bound directly:

```{r sigma2-floor}
s  <- post$sigma2; fl <- params$sigma2_floor
pct1 <- 100 * mean(s < fl * 1.01); pct10 <- 100 * mean(s < fl * 1.10)
cat(sprintf("sigma2: min %.6f  median %.6f  max %.6f   (floor %.4f)\n", min(s), median(s), max(s), fl))
cat(sprintf("  within  1%% of the floor: %5.1f%%\n", pct1))
cat(sprintf("  within 10%% of the floor: %5.1f%%\n", pct10))
cat(sprintf("  distinct values: %d of %d samples\n", length(unique(round(s, 7))), length(s)))
cat(sprintf("  -> %s\n", if (pct1 > 50)
  "PINNED. The BM rate is set by the prior, not the data. Branch-rate statements are unsupported; lower the floor (e.g. dnLoguniform(1e-6, 1)) and rerun."
  else if (pct10 > 5) "close to the floor — watch it, but genuinely estimated"
  else "clear of the floor — genuinely estimated"))
```

```{r ridge, fig.width=10, fig.height=4}
pal_run <- c(`1` = "#E69F00", `2` = "#56B4E9", `3` = "#009E73")
library(patchwork)
# NB: coord_equal is applied to this single plot only, then composed with
# patchwork — never combined with facets.
pa <- ggplot(post, aes(diversification, turnover, colour = run)) +
  geom_point(alpha = 0.2, size = 0.5) +
  scale_colour_manual(values = pal_run) + theme_bw() +
  labs(title = "diversification vs turnover (sampled)")
pb <- ggplot(post, aes(Iteration, sigma2, colour = run)) +
  geom_hline(yintercept = fl, linetype = 2, colour = "red") +
  annotate("text", x = min(post$Iteration), y = fl, label = " prior floor",
           hjust = 0, vjust = -0.6, size = 3, colour = "red") +
  geom_line(alpha = 0.7, linewidth = 0.25) +
  scale_colour_manual(values = pal_run) + theme_bw() +
  labs(title = "sigma2 trace vs its prior floor", y = NULL)
pa | pb
```
