---
title: "Three-clade amphora tree — full analysis (SRVF 38 PC (n-1 max), elastic shape)"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 9
    fig_height: 6
params:
  # This is the SRVF-38 sibling (elastic shape metric). Files carry _srvf20PC.
  # and title. Edit both if you change the analysis.
  run_dir: "/Users/armandleroi/Documents/Amphora_phylogeny/phylogenies/2026_07_three_clade_tree/runs/srvf38pc/output"
  repo_dir: "/Users/armandleroi/Documents/Amphora_phylogeny/phylogenies/2026_07_three_clade_tree"
  data_dir: "/Users/armandleroi/Documents/Amphora_phylogeny/Making amphora types/05_2026_amphora_297_types"
  tree_file: "simple_run_srvf38PC_mcc.tree"
  log_file: "simple_run_srvf38PC.log"
  trees_file: "simple_run_srvf38PC.trees"
  pc_file: "data/mini_character_srvf_38.nex"
  taxa_file: "data/mini_taxa.tsv"
  selected_file: "selected_taxa.csv"
  profile_file: "2026_amphora_297_types_profiles_with_metadata.csv"
  bp_zero: 1950
  pot_offset: 0.03
  pot_height: 0.85
  pot_stretch: 1.0
  fig_w: 10
  fig_h: 12
  burnin: 0.10
  hpd_thin: 3            # subsample posterior trees for the absolute-HPD panel (speed)
  support_cutoff: 0.50
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(ape); library(treeio); library(ggtree); library(ggplot2)
library(dplyr); library(tidyr); library(patchwork)
```

## 1. Read the tree

```{r read}
tr  <- read.beast(file.path(params$run_dir, params$tree_file))
phy <- as.phylo(tr)
tr
```

## 2. Tip metadata

Clade and origin come from `selected_taxa.csv`, **not** the tip label: panathenaics
are labelled `amphora`, so the label's shape token would misclassify them. Year is
the midpoint of each type's date range.

```{r meta}
sel <- read.csv(file.path(params$repo_dir, params$selected_file), stringsAsFactors = FALSE)
meta <- tibble::tibble(label = phy$tip.label) |>
  left_join(sel |> transmute(label = type,
              clade  = dplyr::recode(sub("_(greek|ital)$", "", group),
                        loutro = "loutrophoros", pan = "panathenaic",
                        outgroup = "neck (outgroup)"),
              origin = ifelse(place == "greek_italiote", "Italiote", "Greek"),
              year   = round((date_earliest + date_latest) / 2)),
            by = "label") |>
  mutate(id = sub(".*_([0-9]+)$", "\\1", label),
         tip_label = sprintf("%s %s %d BCE (%s)", origin, clade, -year, id))
stopifnot(!any(is.na(meta$clade)))
pal_clade <- c(pelike = "#E69F00", loutrophoros = "#009E73",
               panathenaic = "#56B4E9", `neck (outgroup)` = "#999999")
pal_origin <- c(Greek = "grey20", Italiote = "#D55E00")
knitr::kable(meta |> count(clade, origin) |> tidyr::pivot_wider(names_from = origin, values_from = n))
```

## 3. Time axis (absolute chronology)

MCC branch lengths are *relative* (youngest tip at 0). Ages in `mini_taxa.tsv` are
BP relative to 1950. Tip ages were estimated, so we anchor on the posterior median
of the youngest tip, then read the axis in calendar years.

```{r tip-ages}
tax <- read.delim(file.path(params$repo_dir, params$taxa_file))
post_all <- read.table(file.path(params$run_dir, params$log_file), header = TRUE,
                       sep = "\t", check.names = FALSE)
keep_rows <- (round(nrow(post_all) * params$burnin) + 1):nrow(post_all)
post_all  <- post_all[keep_rows, ]
tcols <- grep("^t\\[", names(post_all))
tmed  <- sapply(post_all[tcols], median)

# RevBayes orders t[i] by taxon name; verify every median sits inside its prior.
tip_age <- tibble::tibble(label = sort(tax$taxon), age_bp = as.numeric(tmed)) |>
  left_join(tax, by = c(label = "taxon")) |>
  mutate(in_prior = age_bp >= min_age & age_bp <= max_age)
stopifnot(all(tip_age$in_prior))
```

```{r axis}
depth       <- ape::node.depth.edgelength(phy)
root_offset <- max(depth)
rel_age     <- root_offset - depth
anchor_bp   <- min(tip_age$age_bp)                    # youngest tip
node_year   <- (params$bp_zero - anchor_bp) - rel_age
root_year   <- node_year[ape::Ntip(phy) + 1]
cat(sprintf("Youngest tip: %.0f BP = %.0f BCE\n", anchor_bp, params$bp_zero - anchor_bp))
cat(sprintf("Root: %.0f BP = %.0f BCE   (tree height %.0f yr)\n",
            anchor_bp + root_offset, -root_year, root_offset))

p <- ggtree(tr, size = 0.6) %<+% meta
time_axis <- function(headroom = 0.55, left = 0) {
  scale_x_continuous(
    limits = c(min(0, left), root_offset * (1 + headroom)),
    breaks = scales::breaks_pretty(6)(c(min(0, left), root_offset)),
    labels = function(x) { yr <- round(x + root_year)
      ifelse(yr < 0, paste0(abs(yr), " BCE"), ifelse(yr == 0, "0", paste0(yr, " CE"))) })
}
```

## 4. Plot 1 — base tree with support, coloured by clade & origin

```{r plot1, fig.height=11}
p1 <- p +
  geom_tiplab(aes(label = tip_label, colour = clade),
              size = 2.7, offset = 8, show.legend = FALSE) +
  geom_tippoint(aes(colour = clade, shape = origin), size = 2.4) +
  geom_text2(aes(subset = !isTip & !is.na(posterior) &
                          posterior >= params$support_cutoff,
                 label = sprintf("%.2f", posterior)),
             hjust = 1.2, vjust = -0.5, size = 2.5, colour = "grey30") +
  scale_colour_manual(values = pal_clade, name = "Clade") +
  scale_shape_manual(values = c(Greek = 16, Italiote = 17), name = "Origin") +
  theme_tree2() + coord_cartesian(clip = "off") +
  labs(x = NULL, title = "MCC tree, node labels = posterior probability") +
  time_axis()
p1
```

## 5. Plot 2 — 95% HPD on absolute node ages

The MCC `age_0.95_HPD` is on the relative scale. We recompute absolute ages per
posterior sample — `absolute_age = relative_age(sample) + youngest_tip_age(sample)`
— and take the HPD of that, which propagates anchor uncertainty. With 39 tips the
youngest tip is not unique across samples, so we anchor on the minimum tip age in
each sample rather than a fixed tip. Trees are thinned by `hpd_thin` for speed.

```{r abs-hpd}
TT <- read.table(file.path(params$run_dir, params$trees_file), header = TRUE,
                 sep = "\t", stringsAsFactors = FALSE, quote = "",
                 comment.char = "", check.names = FALSE)
stopifnot(identical(TT$Iteration, read.table(
  file.path(params$run_dir, params$log_file), header = TRUE, sep = "\t",
  check.names = FALSE)$Iteration))
TT <- TT[keep_rows, ]

# thin trees + the aligned t[] matrix identically
sub      <- seq(1, nrow(TT), by = params$hpd_thin)
t_young  <- apply(post_all[tcols], 1, min)[sub]        # youngest tip age per sample
trees    <- read.tree(text = gsub("\\[[^]]*\\]", "", TT$Tree[sub]))
canon    <- sort(phy$tip.label)
ckey     <- function(t, ix) paste(sort(match(t$tip.label[ix], canon)), collapse = ",")

acc <- new.env(hash = TRUE, parent = emptyenv())
for (s in seq_along(trees)) {
  ts <- trees[[s]]; dd <- node.depth.edgelength(ts); pp <- prop.part(ts)
  ages <- (max(dd) - dd)[ape::Ntip(ts) + seq_along(pp)] + t_young[s]
  for (i in seq_along(pp)) { k <- ckey(ts, pp[[i]]); acc[[k]] <- c(acc[[k]], ages[i]) }
}
hdi <- function(x, prob = 0.95) { x <- sort(x); n <- length(x); m <- max(1, floor(n * prob))
  i <- which.min(x[m:n] - x[1:(n - m + 1)]); c(x[i], x[i + m - 1]) }

ntip   <- ape::Ntip(phy); mcc_pp <- prop.part(phy)
node_hpd <- lapply(seq_along(mcc_pp), function(i) {
  a <- acc[[ckey(phy, mcc_pp[[i]])]]
  if (is.null(a)) return(NULL)
  tibble::tibble(node = ntip + i, freq = length(a) / length(trees),
                 med = median(a), lo = hdi(a)[1], hi = hdi(a)[2]) }) |> bind_rows()
tip_hpd <- lapply(seq_len(ntip), function(i) {
  a <- post_all[[tcols[match(phy$tip.label[i], sort(tax$taxon))]]]
  tibble::tibble(node = i, freq = 1, med = median(a), lo = hdi(a)[1], hi = hdi(a)[2]) }) |>
  bind_rows()
age_hpd <- bind_rows(tip_hpd, node_hpd)
```

```{r plot2, fig.height=11}
bp_to_x <- function(bp) (params$bp_zero - bp) - root_year
bars <- age_hpd |> left_join(p$data |> select(node, y), by = "node") |>
  mutate(x_lo = bp_to_x(hi), x_hi = bp_to_x(lo), x_med = bp_to_x(med),
         kind = ifelse(node <= ntip, "Tip age", "Node age"))
p2 <- p +
  geom_segment(data = bars, aes(x = x_lo, xend = x_hi, y = y, yend = y, colour = kind),
               linewidth = 2, alpha = 0.4, lineend = "round") +
  geom_point(data = bars, aes(x = x_med, y = y), size = 0.8, colour = "grey20") +
  geom_tiplab(aes(label = tip_label), size = 2.5, offset = 8) +
  scale_colour_manual(values = c(`Node age` = "#377EB8", `Tip age` = "#4DAF4A"),
                      name = "95% HPD") +
  theme_tree2() + coord_cartesian(clip = "off") +
  labs(title = "95% HPD on absolute ages (anchor uncertainty propagated)",
       subtitle = "grey dots = posterior median; freq column = clade frequency in posterior") +
  time_axis(headroom = 0.75, left = min(bars$x_lo))
p2
```

Node bars are wide because within-clade support is low: a node present in only a
fraction of posterior samples (`freq`) has its age estimated from those samples
alone. The deep squat/elongate split is the exception — well supported, tight bar.

## 6. Plot 3 — support as node size, sampled-ancestor probability

```{r plot3, fig.height=11}
p3 <- p +
  geom_nodepoint(aes(size = posterior), colour = "grey25", alpha = 0.7) +
  geom_tippoint(aes(fill = sampled_ancestor), shape = 21, size = 3, colour = "grey20") +
  geom_tiplab(aes(label = tip_label), size = 2.5, offset = 8) +
  scale_size_continuous(range = c(0.5, 5), name = "Posterior") +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "P(sampled\nancestor)") +
  theme_tree2() + coord_cartesian(clip = "off") +
  labs(title = "Clade support and sampled-ancestor probability") +
  time_axis()
p3
```

## 7. Plot 4 — pot profiles at the tips (filled by clade)

```{r profiles}
prof_all <- read.csv(file.path(params$data_dir, params$profile_file))
stopifnot(all(phy$tip.label %in% prof_all$type))
tipxy <- p$data |> filter(isTip) |> select(label, tx = x, ty = y)
headroom <- 0.75
k  <- (root_offset * (1 + headroom) / ape::Ntip(phy)) * (params$fig_h / params$fig_w)
sy <- params$pot_height / 3; sx <- sy * k * params$pot_stretch
pots <- prof_all |> filter(type %in% phy$tip.label) |>
  select(label = type, point_order, px = x, py = y) |>
  left_join(tipxy, by = "label") |> left_join(select(meta, label, clade), by = "label") |>
  arrange(label, point_order) |>
  mutate(x = tx + params$pot_offset * root_offset + px * sx, y = ty + py * sy)
pot_span <- max(abs(prof_all$x)) * sx
```

```{r plot4, fig.width=params$fig_w, fig.height=params$fig_h}
p4 <- p +
  geom_polygon(data = pots, aes(x = x, y = y, group = label, fill = clade),
               colour = "grey25", linewidth = 0.22) +
  geom_tippoint(size = 1.1, colour = "grey30") +
  geom_text2(aes(subset = !isTip & !is.na(posterior) & posterior >= params$support_cutoff,
                 label = sprintf("%.2f", posterior)), hjust = 1.2, vjust = -0.5,
             size = 2.4, colour = "grey30") +
  geom_text(data = distinct(pots, label, tx, ty) |> left_join(meta, by = "label"),
            aes(x = tx + params$pot_offset * root_offset + 1.25 * pot_span, y = ty,
                label = sprintf("%s, %d BCE", origin, -year)),
            hjust = 0, size = 2.5, colour = "grey20") +
  scale_fill_manual(values = pal_clade, name = "Clade") +
  theme_tree2() + coord_cartesian(clip = "off") +
  labs(title = "MCC tree with pot profiles at the tips") +
  time_axis(headroom = headroom)
p4
```

## 8. Plot 5 — pots filled by measured shape (PC1), outlined by clade

These are the same PCs the model was fitted to (`r params$pc_file`).

```{r read-pcs}
parse_cont_nex <- function(f) {
  L <- readLines(f); L <- L[grepl("^[[:space:]]+[^[:space:]]+[[:space:]]+-?[0-9]", L)]
  parts <- strsplit(trimws(L), "[[:space:]]+"); tx <- vapply(parts, `[`, "", 1)
  vals <- lapply(parts, function(p) as.numeric(p[-1])); ord <- unique(tx)
  out <- t(vapply(ord, function(t) unlist(vals[tx == t]),
                  numeric(length(unlist(vals[tx == ord[1]])))))
  colnames(out) <- paste0("PC", seq_len(ncol(out))); out
}
pcs  <- parse_cont_nex(file.path(params$repo_dir, params$pc_file))
npc  <- min(5, ncol(pcs))
pcdf <- tibble::as_tibble(pcs[, 1:npc], rownames = "label")
stopifnot(all(phy$tip.label %in% pcdf$label))
aspect <- sapply(pcdf$label, function(t) {
  s <- filter(prof_all, type == t); (max(s$y) - min(s$y)) / (max(s$x) - min(s$x)) })
r_pc1 <- cor(pcdf$PC1, aspect)
cat(sprintf("cor(PC1, height/width) = %.2f  (|r| near 1 => PC1 is the squat/slender axis)\n", r_pc1))
# orient so that positive = slender (taller), for a consistent colour scale
if (r_pc1 < 0) { pcs[, "PC1"] <- -pcs[, "PC1"]; pcdf$PC1 <- -pcdf$PC1 }
```

```{r plot5, fig.width=params$fig_w, fig.height=params$fig_h}
pots_pc <- pots |> left_join(pcdf |> select(label, PC1), by = "label")
p5 <- p +
  geom_polygon(data = pots_pc, aes(x = x, y = y, group = label, fill = PC1, colour = clade),
               linewidth = 0.5) +
  geom_text(data = distinct(pots, label, tx, ty) |> left_join(meta, by = "label"),
            aes(x = tx + params$pot_offset * root_offset + 1.25 * pot_span, y = ty,
                label = sprintf("%s, %d BCE", origin, -year)),
            hjust = 0, size = 2.5, colour = "grey20") +
  scale_fill_gradient2(low = "#56B4E9", mid = "grey93", high = "#E69F00", midpoint = 0,
                       name = "PC1\n(squat → slender)") +
  scale_colour_manual(values = pal_clade, name = "Clade") +
  guides(colour = guide_legend(override.aes = list(fill = NA, linewidth = 1.2))) +
  theme_tree2() + coord_cartesian(clip = "off") +
  labs(title = "Fill = measured shape (PC1); outline = clade") +
  time_axis(headroom = headroom)
p5
```

```{r morphospace, fig.width=7.5, fig.height=6}
msp <- pcdf |> left_join(meta, by = "label")
ggplot(msp, aes(PC1, PC2)) +
  geom_hline(yintercept = 0, colour = "grey85") + geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(aes(fill = clade, shape = origin), size = 3.5, colour = "grey25", stroke = 0.4) +
  ggrepel::geom_text_repel(aes(label = sprintf("%s %d BCE", id, -year)),
                           size = 2.4, colour = "grey35", max.overlaps = 25) +
  scale_fill_manual(values = pal_clade, name = "Clade") +
  scale_shape_manual(values = c(Greek = 21, Italiote = 24), name = "Origin") +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  theme_bw() +
  labs(title = "Morphospace of the 39 taxa",
       subtitle = "PC1 = squat↔slender; clades overlap heavily (why within-clade support is low)")
```

## 9. Posterior parameter estimates & traces

```{r params-post}
logs <- list.files(params$run_dir, pattern = "_run_[0-9]+\\.log$", full.names = TRUE)
post <- lapply(seq_along(logs), function(i) {
  d <- read.table(logs[i], header = TRUE, sep = "\t", check.names = FALSE)
  d <- d[-seq_len(ceiling(nrow(d) * params$burnin)), ]; d$run <- factor(i); d }) |> bind_rows()
pars <- c("speciation_rate", "extinction_rate", "psi", "sigma2", "origin_time")
post |> select(all_of(pars)) |> pivot_longer(everything()) |> group_by(name) |>
  summarise(median = median(value), lower = quantile(value, .025),
            upper = quantile(value, .975)) |> knitr::kable(digits = 5)
```

```{r trace, fig.height=5}
post |> select(Iteration, run, all_of(pars)) |> pivot_longer(all_of(pars)) |>
  ggplot(aes(Iteration, value, colour = run)) +
  geom_line(alpha = 0.6, linewidth = 0.25) +
  facet_wrap(~name, scales = "free_y") +
  scale_colour_brewer(palette = "Dark2") + theme_bw() +
  labs(title = "Traces across the three runs", y = NULL)
```

## 10. MCMC diagnostics

```{r ess}
library(coda)
runs <- split(post, post$run)
allp <- c(pars)
ess  <- sapply(runs, function(d) sapply(d[allp], \(x) effectiveSize(as.mcmc(x))))
psrf <- gelman.diag(mcmc.list(lapply(runs, \(d) as.mcmc(as.matrix(d[allp])))),
                    multivariate = FALSE, autoburnin = FALSE)$psrf
tibble::tibble(parameter = allp, run1 = round(ess[, 1]), run2 = round(ess[, 2]),
               run3 = round(ess[, 3]), psrf = round(psrf[allp, 1], 3)) |>
  arrange(pmin(run1, run2, run3)) |> knitr::kable()
```

`speciation_rate`/`extinction_rate` are non-identifiable (FBD ridge); reparameterise
and the informed directions sample perfectly. `sigma2` is the number to watch across
PC counts — it slides toward its 0.001 floor as low-variance PCs are added.

```{r reparam}
post <- post |> mutate(net_div = speciation_rate - extinction_rate,
                       turnover = extinction_rate / speciation_rate)
cat(sprintf("cor(lambda, mu) = %.4f\n", cor(post$speciation_rate, post$extinction_rate)))
post |> select(net_div, turnover, sigma2) |> pivot_longer(everything()) |> group_by(name) |>
  summarise(median = median(value), lower = quantile(value, .025),
            upper = quantile(value, .975)) |> knitr::kable(digits = 5)
```

```{r ridge, fig.width=10, fig.height=4}
pal_run <- c(`1` = "#E69F00", `2` = "#56B4E9", `3` = "#009E73")
(ggplot(post, aes(speciation_rate, extinction_rate, colour = run)) +
   geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = 2) +
   geom_point(alpha = 0.25, size = 0.6) + coord_equal() +
   scale_colour_manual(values = pal_run) + theme_bw() +
   labs(title = "lambda vs mu: a line, not a cloud")) |
(ggplot(post, aes(Iteration, sigma2, colour = run)) +
   geom_line(alpha = 0.7, linewidth = 0.25) +
   scale_colour_manual(values = pal_run) + theme_bw() +
   labs(title = "sigma2 trace (BM rate)", y = NULL))
```
