---
title: "Three-clade amphora tree — visualised"
subtitle: "dated MCC tree: pot profiles, clade/origin colouring, posterior support, age HPDs"
output:
  html_document:
    toc: true
    toc_float: true
params:
  run_dir: "/Users/armandleroi/Documents/Amphora_phylogeny/three_clade_tree/hpc_output"
  data_dir: "/Users/armandleroi/Documents/Amphora_phylogeny/2026_amphora_297_types"
  taxa_dir: "/Users/armandleroi/Documents/Amphora_phylogeny/three_clade_tree"
  support_cutoff: 0.5     # only show well-supported node posteriors
  fig_w: 10
  fig_h: 12
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = params$fig_w, fig.height = params$fig_h)
library(treeio); library(ggtree); library(ape); library(ggplot2); library(dplyr)
```

```{r base}
trb <- read.beast(file.path(params$run_dir, "simple_run_mcc.tree"))
phy <- as.phylo(trb)
ntip <- Ntip(phy)

sel  <- read.csv(file.path(params$taxa_dir, "selected_taxa.csv"), stringsAsFactors = FALSE)
meta <- sel |>
  transmute(label = type,
            clade  = dplyr::recode(sub("_(greek|ital)$", "", group),
                                   loutro = "loutrophoros", pan = "panathenaic",
                                   outgroup = "neck (outgroup)"),
            origin = ifelse(place == "greek_italiote", "Italiote", "Greek"))

# absolute chronology: anchor node depths on tip-age midpoints (years BP -> BCE)
ages  <- read.delim(file.path(params$taxa_dir, "data/mini_taxa.tsv"))
tipBP <- setNames((ages$min_age + ages$max_age) / 2, ages$taxon)
depth <- node.depth.edgelength(phy)
rootBP <- median(tipBP[phy$tip.label] + depth[1:ntip])
nodeYr <- 1950 - (rootBP - depth)          # calendar year, negative = BCE
root_yr <- nodeYr[ntip + 1]
present_yr <- max(nodeYr[1:ntip])          # youngest tip's calendar year

pal_clade <- c(pelike = "#E69F00", loutrophoros = "#009E73",
               panathenaic = "#56B4E9", `neck (outgroup)` = "#999999")
pal_origin <- c(Greek = "grey20", Italiote = "#D55E00")
```

Root ≈ **`r sprintf('%.0f BCE', -root_yr)`**, youngest tip ≈ **`r sprintf('%.0f BCE', -present_yr)`**.
Deepest split (posterior 1.00) separates **squat (pelike)** from **elongate
(loutrophoros + panathenaic)**; Greek and Italiote intermingle within every clade.

## 1. Dated tree with pot profiles

Pots filled by **clade**, tip labels coloured by **origin**, posterior support
(≥ `r params$support_cutoff`) at nodes, calendar axis.

```{r pots}
p <- ggtree(trb) + theme_tree2()
xr <- max(p$data$x)
tipxy <- p$data |> filter(isTip) |> transmute(label, tx = x, ty = y)

prof <- read.csv(file.path(params$data_dir,
                 "2026_amphora_297_types_profiles_with_metadata.csv"),
                 stringsAsFactors = FALSE)
xt <- max(tipxy$tx); gap <- 0.05 * xr
k_asp <- (xr * 1.9 / ntip) * (params$fig_h / params$fig_w)
sy <- 0.85 / 3; sx <- sy * k_asp
pots <- prof |>
  filter(type %in% phy$tip.label) |>
  transmute(label = type, point_order, px = x, py = y) |>
  left_join(tipxy, by = "label") |> left_join(meta, by = "label") |>
  arrange(label, point_order) |>
  mutate(x = xt + gap + px * sx, y = ty + py * sy)
lab <- meta |> left_join(tipxy, by = "label") |>
  mutate(x = xt + gap + 2.4 * max(abs(prof$x)) * sx,
         txt = sprintf("%s  %s", origin, clade))

brks_yr <- seq(-1200, 0, by = 200); brks_x <- brks_yr - root_yr
keep <- brks_x >= 0 & brks_x <= xr

p +
  geom_tippoint(aes(colour = origin),
                data = left_join(p$data, meta, by = "label"), size = 1.7) +
  geom_polygon(data = pots, aes(x, y, group = label, fill = clade),
               colour = "grey25", linewidth = 0.2) +
  geom_text(data = lab, aes(x, ty, label = txt, colour = origin),
            hjust = 0, size = 2.6) +
  geom_text2(aes(subset = !isTip & !is.na(as.numeric(posterior)) &
                          as.numeric(posterior) >= params$support_cutoff,
                 label = sprintf("%.2f", as.numeric(posterior))),
             hjust = 1.2, vjust = -0.4, size = 2.3, colour = "grey35") +
  scale_fill_manual(values = pal_clade, name = "Clade") +
  scale_colour_manual(values = pal_origin, name = "Origin") +
  scale_x_continuous(breaks = brks_x[keep], labels = paste0(-brks_yr[keep], " BCE")) +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(6, 170, 6, 6), axis.text.x = element_text(size = 8)) +
  labs(title = "Dated MCC tree with pot profiles",
       subtitle = sprintf("root %.0f BCE · fill = clade · label colour = origin", -root_yr))
```

## 2. Node ages with 95% HPD bars

Blue bars = 95% highest-posterior-density interval on each node's age. Tips
coloured by origin, posterior support at nodes. This shows how well the dates are
pinned down (short bar = confident, long bar = uncertain).

```{r hpd, fig.height=11}
ph <- revts(ggtree(trb) %<+% meta + theme_tree2())   # x = age before youngest tip (<= 0)
ph +
  geom_range("age_0.95_HPD", color = "#377EB8", alpha = 0.4, size = 2) +
  geom_tippoint(aes(colour = origin), size = 1.6) +
  geom_tiplab(aes(colour = origin, label = paste(origin, clade)),
              size = 2.4, hjust = -0.02) +
  geom_text2(aes(subset = !isTip & !is.na(as.numeric(posterior)) &
                          as.numeric(posterior) >= params$support_cutoff,
                 label = sprintf("%.2f", as.numeric(posterior))),
             hjust = 1.2, vjust = -0.4, size = 2.3, colour = "grey35") +
  scale_colour_manual(values = pal_origin, name = "Origin", na.value = "grey50") +
  scale_x_continuous(breaks = seq(-1200, 0, 200) - present_yr,
                     labels = paste0(seq(1200, 0, -200), " BCE")) +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(6, 140, 6, 6), axis.text.x = element_text(size = 8)) +
  labs(title = "Node ages with 95% HPD intervals",
       subtitle = sprintf("root %.0f BCE · blue bar = 95%% HPD on node age", -root_yr))
```

## 3. Membership of the two deep clades

```{r members}
rc <- phy$edge[phy$edge[, 1] == (ntip + 1), 2]
ingroup <- rc[rc > ntip]
ch <- phy$edge[phy$edge[, 1] == ingroup, 2]
for (i in seq_along(ch)) {
  tips <- if (ch[i] <= ntip) phy$tip.label[ch[i]] else extract.clade(phy, ch[i])$tip.label
  cat(sprintf("deep clade %d (%d tips):\n", i, length(tips)))
  print(table(meta$clade[match(tips, meta$label)]))
  cat("\n")
}
```
